---
title: "fec16"
author: "Irene Ryan, Marium Tapal, and Rana Gahwagy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fec16}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4.5
)
library(fec16)
library(scales)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
```

The `fec16` package provides tamed datasets of the Federal Election Commission (FEC)'s 2015-2016 election cycle. 

## Inspiration 

We wanted to create a data package that is easy to use for people just beginning to learn R, for example, in introductory statistics and data science classes. For this purpose, this package is nice because students and instructors do not need to worry about the unnecessary data wrangling and can immediately use the data for analysis. 

Our package is inspired by [Hadley Wickham's](https://github.com/hadley/nycflights13) `nycflights13` package. We used the data taming principles from [Albert Kim's](https://github.com/rudeboybert/fivethirtyeight) `fivethirtyeight` package, which are explained in detail [here](https://escholarship.org/uc/item/0rx1231m).


### Basics of tame data 

We used the tame data principles to produce uniform data frames that are easy to navigate and link together. The following are the guidelines we used to tame our data:

1. Variable names: 
   * Lower case and with underscores instead of spaces using `clean_names()`
   * 20 characters or less
   * The same names were used for the same variables in different data frames 
   * Names to be meaningful 
2. Variable types:
   * Encode variable with dates with `as.Date`, unless it is only a year then it would be numeric type
   * Encode categorical variables that a have a limited number of values with `as.factor` otherwise it would be of character type
3. Tidy or long data format instead of wide 
4. Missing data has value `NA`
5. Removed signs such as % and $ attached to numeric values

## Other notes

-  We dropped variables that are not useful for analysis from the original FEC datasets 
-  We summarized large datasets with relevant data in order to compress its size

## Data frames included: 

All the datasets are taken from [FEC](https://www.fec.gov/data/browse-data/?tab=bulk-data) 

`candidates`: all candidates registered with the FEC during the 2015-2016 election cycle

`committees`: all committees registered with the FEC during the 2015-2016 election cycle

`results_house`: the results of the 2016 Congressional elections

`results_senate`: the senate results of the 2016 general presidential election

`results_president`: the final results of the 2016 general presidential election

`individuals`: a sample of 5000 individual contributions to candidates/committees during the primary and general 2016 elections

`committee_contributions`: total contributions, aggregated by candidate, from committees

`campaigns`: summary of financial information of the House/Senate Current Campaigns

## Who should use this package? 

Mainly, anyone interested in US politics and elections and wants to use actual data to think critically and make inferences about it should use this package. Additionally, since we made this package with students and their instructors in mind to make working with the data itself as smooth as possible, it is a good set of data for teaching. Moreover, this package can provide inspiration for questions and projects that are data-driven and based on the real world. 

## What does the data look like?

The first six rows of the `results_house` dataset look like:
```{r}
head(results_house)
```

## What can we do with this data?

We can use this package to address the (non-exhaustive) list of questions: 

  1. Which presidential candidate won majority in more states?
  2. What is the relationship between contributions of candidates and total votes they get? 
  3. Which candidate got the most popular vote and how many?
  4. Is there a trend in the contributions to a candidate depending on their party?

To answer our questions we can make use of some **data wrangling** and **data visualization** techniques. Some examples (addressing the questions above) are shown below:

### Example 1:

**Which presidential candidate won a majority in more states, and how many votes did they get?**

Using the `results_president` dataset, we can also see what actually happened in the 2016 elections.

Here is how we can summarize the number of wins by candidate:
```{r}
wins <- results_president %>%
  group_by(cand_id, name_last) %>%
  summarise(
    total_votes = sum(general_votes, na.rm = TRUE),
    states_won = sum(won)
  ) %>%
  arrange(desc(total_votes))
wins
```

Compare these results to [those reported by the Wikipedia](https://en.wikipedia.org/wiki/2016_United_States_presidential_election).
We can show the results using a simple bar chart:

Each win is for a single state. There are 51 total wins for the 50 States and Washington, D.C.
```{r}
ggplot(wins %>% head(6), aes(x = reorder(name_last, total_votes), y = total_votes, fill = cand_id)) + 
  geom_col() +
  scale_fill_discrete(guide = FALSE) + 
  labs(
    title = "2016 Presidential Election",
    subtitle = "Total votes",
    x = NULL, y = "Number of Votes"
  ) + 
  coord_flip()
```

We can see that Hillary Clinton earned the most votes. 

### Example 2:

**What is the relationship between contributions of candidates and total votes they get?**

Here we investigate what kind of relationship, if any, the candidates with over a 1000 votes got with the total contributions they made. We are interested in candidates with an ID number so we can join them with the contribution data set by using `inner_join`.

```{r, eval=TRUE}
results_by_cand <- results_house %>%
  group_by(cand_id, name_last) %>%
  summarise(sum_votes = sum(general_votes)) %>%
  filter(sum_votes > 1000) %>%
  inner_join(contributions, by = "cand_id")
results_by_cand
```

Next, we plot the contributions and votes on a scatter plot and plot a trend line that would make it easy for us to see the relationship. Since there are outliers in the data, it is best to not use a continuous axis scale in order to see all of the points. 

```{r eval=FALSE, message=FALSE}
ggplot(results_by_cand, mapping = aes(x = total_contributions, y = sum_votes)) +
  geom_point() +
  scale_x_log10(labels = comma) +
  scale_y_sqrt(labels = comma) +
  geom_smooth(method = "auto") +
  labs(title = "Contributions vs. Votes in 2016", 
       x = "Contributions in US Dollars", y = "Total Votes") + 
  geom_text_repel(aes(label = last_name))
```

As we can see, the highest contributors got the highest amount of votes so it has a positive correlation. 


### Example 3:

**What is the distribution of number of votes across winners in the US House of Representatives?**

Visualize the results of the popular vote in the elections and see how many people voted: 
```{r message=FALSE, warning=FALSE}
house_winners <- results_house %>%
  filter(won, state %in% state.abb) %>%
  group_by(cand_id, name_last) %>%
  summarize(
    total_votes = sum(general_votes),
    total_pct = sum(general_percent),
    party = first(party)
  ) %>%
  filter(total_pct <= 1)
  
ggplot(house_winners, 
       aes(x = total_votes, y = total_pct, 
           color = party)) +
  geom_point() +
  labs(
    title = "Not all Congressional Races are the same", 
    fill = "Candidate") +
  scale_y_continuous(labels = comma)
```




### Example 4:

**Is there a trend in the contributions to a candidate depending on their party?**

We can create a scatter-plot of total contributions per candidate, colored by party (only including Democratic and Republican parties).
```{r message = FALSE, eval=FALSE}
joined_data <- candidates %>%
  full_join(committee_contributions, by = "cand_id") %>%
  filter(!is.na(total_contributions)) %>%
  filter(cand_pty_aff == "REP" | cand_pty_aff == "DEM") %>%
  filter(total_contributions <= 1e+07)

ggplot(joined_data, aes(x = cand_id, y = total_contributions, color = cand_pty_aff)) +
  geom_point() +
  geom_jitter() +
  theme(axis.text.x = element_blank()) +
  labs(
    title = "Contributions per Candidate by Party",
    x = "Candidates", y = "Total Contributions ($)",
    color = "Candidate Party Affiliation"
  ) +
  scale_y_continuous(labels = comma)
```

It doesn't look like there are any specific patterns in the contributions (for contributions under $ 10 million).


